---
title: "Final Project Draft of Code"
author: "Third Time's the Charm"
date: "2025-10-23"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 0: Why?


**Dataset Chosen:**  
Hotel Reservations (~35,000 rows)  
Predicts hotel booking cancellations and potential no-shows. The business use case focuses on **optimizing overbooking decisions** to maximize hotel occupancy and revenue.

### Scenario: Meet Ron, the Hotel Manager

Ron is the general manager of a mid-sized city hotel that operates near full capacity during most of the year. Recently, Ron has been facing a major operational headache: **cancellations and overbookings are wrecking his occupancy numbers**. Some nights he ends up with 20 empty rooms because too many guests canceled at the last minute. Other nights, he slightly overbooks to compensate—but the wrong guests show up, forcing him to **walk** customers to nearby hotels, costing him money and damaging his reputation.

Ron's front desk staff is frustrated, customer complaints are rising, and corporate has asked him to “tighten up” revenue performance. What Ron needs is a **predictive system**—something that can tell him, *“Which bookings are most likely to cancel?”* so he can safely overbook when appropriate, avoid walking guests, and recover revenue that would otherwise be lost.

This is the exact business problem our project solves.


### Context (from Kaggle)
> “The online hotel reservation channels have dramatically changed booking possibilities and customers’ behavior. A significant number of hotel reservations are called-off due to cancellations or no-shows. The typical reasons for cancellations include change of plans, scheduling conflicts, etc. This is often made easier by the option to do so free of charge or preferably at a low cost which is beneficial to hotel guests but it is a less desirable and possibly revenue-diminishing factor for hotels to deal with.”

### Business Use Case
We are building a model to predict which hotel reservations are most likely to cancel.  
Although cancellations often result from a change of plans or scheduling conflicts, they impose a direct financial cost on the hotel.  
To mitigate these losses and **maintain 100% occupancy**, the hotel can strategically **overbook** based on model predictions of cancellation likelihood.

**Main Question:**  
Can we predict which hotel reservations are most likely to be canceled, and use these predictions to make more profitable overbooking decisions?

**Why it’s answerable with this dataset:**  
- The dataset includes **booking-level data** for over 35 000 reservations with both categorical and numerical features (lead time, room type, guest history, etc.).  
- The dependent variable `booking_status` is binary (Canceled / Not Canceled), making it well-suited for **classification via logistic regression**.  
- With exploratory cleaning and modeling, we can estimate how much predictive signal exists and whether cancellation behavior is predictable enough to support overbooking decisions.


### Financial Assumptions
- **Revenue per booked room:** \$250 per night on average price in a city
(source: https://www.coohom.com/article/how-much-do-hotel-rooms-cost)
- **Cancellations:** Guests can cancel free of charge at any time; each cancellation results in a \$250 revenue loss (from the cost of the room)  
- **Overbooking cost:** If the hotel overbooks and cannot accommodate a guest (because no cancellations occur), it incurs \$500 per night because the cost of walking a guest is ~2x the room cost in re-accommodation, transportation, goodwill, and/or compensation costs.
(souce: https://ecornell-impact.cornell.edu/the-cheapest-and-best-approach-to-overbooking/)
- **Variable costs per room:** Considered sunk and excluded from analysis (the cost of turning a room is identical regardless of occupancy).  
- **No difference** between cancellation and no-show costs.


### Cost–Benefit Matrix

| Case | Model Prediction | Actual Outcome | Financial Impact |
|------|--------------|----------------|-----------------|
| **True Positive (TP)** | "Will Cancel" | Guest cancels | **+ \$250** (revenue from replacement guest) |
| **True Negative (TN)** | "Will Not Cancel" | Guest does not cancel | **\$0** (no change in revenue) |
| **False Negative (FN)** | "Will Not Cancel" | Guest cancels | **– \$250** (lost booking revenue) |
| **False Positive (FP)** | "Will Cancel" | Guest shows up | **– \$500** (overbooking penalty) |


### Objective
Develop a predictive model that helps hotels **balance the trade-off** between overbooking risk and lost revenue due to cancellations, thereby maximizing **expected net revenue** and improving operational decision-making.


### Step 0.5: Load Packages
```{r}
library(caret)
library(neuralnet)
library(class)
library(C50)
library(ggplot2)
library(randomForest)
library(kernlab)
```


## Step 1: Load Data
```{r}
hotel <- read.csv("Hotel Reservations.csv", stringsAsFactors = TRUE)
```

## Step 2: Clean Data
```{r}
str(hotel)
summary(hotel)

```
Let's remove booking id. Let's also convert our dependent variable (booking status) into 0s and 1s.
```{r}
hotel$Booking_ID = NULL
hotel$booking_status <- ifelse(hotel$booking_status == "Canceled", 1, 0)
str(hotel)
summary(hotel)
```

## Step 3: Split Data
We have 36000 data entries. In order to build a second level model, we are going to keep 50% for evaluation.
Let's dummify, scale, and split the data.
```{r}
hotel_dummies <- as.data.frame(model.matrix(~ . - booking_status - 1, data = hotel))
hotel_dummies$booking_status <- hotel$booking_status

trainprop <- 0.5
set.seed(12345)
trainrows <- sample(1:nrow(hotel_dummies), trainprop * nrow(hotel_dummies))

minmax <- function(x) {
  if (is.numeric(x)) {
    return((x - min(x)) / (max(x) - min(x)))
  } else {
    return(x)
  }
}
hotel_scaled <- as.data.frame(lapply(hotel_dummies, minmax))

hotel_train <- hotel_dummies[trainrows, ]
hotel_test  <- hotel_dummies[-trainrows, ]

hotel_train_scaled <- hotel_scaled[trainrows, ]
hotel_test_scaled <- hotel_scaled[-trainrows, ]
str(hotel_train)
```


## Step 4: Build Models

### Step 4.1: Build LogReg Model

Our baseline model will be a **vanilla Logistic Regression (LogReg)**, which predicts the probability of a booking being canceled.  
This serves as the foundation for comparison with future models (e.g., neural networks, decision trees, etc.).
```{r, cache=TRUE}
logreg_model <- glm(booking_status ~ ., data = hotel_train, family = "binomial")
```

Let's also build an enhanced model. To improve the predictive performance of our baseline logistic regression model, we introduce several **interaction terms**. These interactions allow the model to capture relationships between variables that jointly influence cancellation behavior. Including these terms helps the model represent more complex customer patterns that cannot be captured by individual predictors alone.

#### **1. `lead_time × avg_price_per_room`**
This interaction captures how the impact of lead time changes depending on room price.
- Customers who book **high-priced rooms far in advance** may be more likely to cancel as plans, budgets, or alternatives change.
- Customers booking inexpensive rooms far in advance may behave differently.
This term allows the model to account for situations where **price modifies the effect of lead time** on cancellations.

#### **2. `no_of_special_requests × repeated_guest`**
This interaction differentiates behavior between repeat guests and first-time customers.
- Repeat guests with many special requests may be more committed (e.g., loyal or planned stays).
- First-time guests with many requests may be more likely to cancel if expectations shift.
This term helps the model identify how **customer loyalty interacts with booking complexity**.

#### **3. `no_of_weekend_nights × no_of_week_nights`**
This interaction captures the structure of the guest’s stay.
- Longer, more complex trips (spanning both weekend and weekday nights) may carry higher cancellation risks.
This term helps the model evaluate how the **overall length and nature of the stay** influence cancellation behavior.

#### **4. `arrival_month × market_segment_type` (applied across dummy variables)**
After dummification, `market_segment_type` becomes several binary indicators (e.g., Online, Offline, Corporate).  
Interacting `arrival_month` with each segment dummy allows the model to learn **segment-specific seasonal cancellation patterns**:
- Corporate travelers may cancel more during holiday months.
- Online/OTA bookings often surge—and cancel more—during peak travel seasons or promotional periods.
- Offline bookings (e.g., via agents) may follow entirely different seasonal trends.  
By including these interactions, the model captures how **seasonality modifies cancellation behavior across different booking channels**.


```{r, cache=TRUE}
logreg_model_enhanced <- glm(
  booking_status ~ 
    . +
    lead_time * avg_price_per_room +
    no_of_special_requests * repeated_guest +
    no_of_weekend_nights * no_of_week_nights +
    arrival_month * market_segment_typeComplementary +
    arrival_month * market_segment_typeCorporate +
    arrival_month * market_segment_typeOffline +
    arrival_month * market_segment_typeOnline,
  data = hotel_train,
  family = "binomial"
)


```


### Step 4.2: Build KNN Model
Here, we separate our predictors and target variables into training and testing sets. We will predict and evaluate in step 5. 
```{r}
train_features <- subset(hotel_train_scaled, select = -booking_status)
train_labels <- hotel_train_scaled$booking_status
test_features <- subset(hotel_test_scaled, select = -booking_status)
test_labels <- hotel_test_scaled$booking_status
```

### Step 4.3: Build ANN Model
We are now building an ANN model with a hidden layer of 5 neurons. With structured data, typically a shallow network can capture enough valuable relationships. 
```{r, cache = TRUE}
ann_model <- neuralnet(
  booking_status ~ .,
  data = hotel_train_scaled,
  lifesign = "minimal",
  stepmax = 1e6,
  hidden = 2
)
```


### Step 4.4: Build Decision Tree Model
We will now build a normal decision tree model. We will not impose a cost matrix here but we will in the second level decision tree. 
```{r, cache=TRUE}
m_decision <- C5.0(
  as.factor(booking_status) ~ .,
  data = hotel_train
)

plot(m_decision)
```

The decision tree relies most heavily on a few key predictors. **Lead time** is by far the strongest driver of cancellation behavior, appearing in virtually every major split. The model also heavily uses **number of special requests**, **repeated guest status**, and **number of week nights** to differentiate likely cancellations from reliable bookings. Additionally, **price-related features** (such as `avg_price_per_room`) and **market segment type** help the tree distinguish different types of guests. Overall, the tree prioritizes guest commitment indicators (lead time, repeat status), stay characteristics (week nights, weekend nights), and booking channel/price information to identify cancellation risk.

### Step 4.5: Build SVM
We will train a basic SVM with a RBF kernel.
```{r, cache=TRUE}
hotel_svm_basic <- ksvm(
  as.factor(booking_status) ~ .,
  data = hotel_train_scaled,
  kernel = "rbfdot",
  prob.model = TRUE
)
```

We will also build an enhanced SVM model. We  refine the model by tuning two key hyperparameters:

**C (Regularization parameter):**  
Controls the trade-off between maximizing the margin and minimizing classification errors.

- A **larger C** (e.g., 5) penalizes misclassifications more heavily, producing a **smaller margin** and a tighter, more detailed fit to the data (lower bias, higher variance).  
- A **smaller C** allows more errors but produces a **wider margin**, leading to a smoother, more generalizable model (higher bias, lower variance).

**σ (Sigma in the RBF kernel):**  
Determines how far the influence of each support vector extends.

- A **smaller σ** (e.g., 0.05) makes each support vector’s influence very local, enabling the model to capture **fine-grained, nonlinear boundaries** (risk of overfitting).  
- A **larger σ** smooths the decision boundary, reducing overfitting but potentially missing subtle structure in the data.

This improved SVM aims to better separate likely cancellations from non-cancellations by allowing a more flexible, finely tuned boundary.
```{r, cache=TRUE}
hotel_svm_enhanced <- ksvm(
  as.factor(booking_status) ~ .,
  data = hotel_train_scaled,
  kernel = "rbfdot",
  C = 1,
  kpar = list(sigma = 0.1),
  prob.model = TRUE
)
```



### Step 4.6: Build Random Forest
We now train a Random Forest classifier using the unscaled dataset. We include 500 trees and use the square-root rule for mtry, which is standard for classification tasks.
```{r, cache=TRUE}
hotel_train$booking_status <- as.factor(hotel_train$booking_status)
hotel_test$booking_status  <- as.factor(hotel_test$booking_status)

rf_model <- randomForest(
  x = hotel_train[, -which(names(hotel_train) == "booking_status")],
  y = hotel_train$booking_status,
  ntree = 500,
  mtry = sqrt(ncol(hotel_train) - 1),
  importance = TRUE
)

plot(rf_model)
varImpPlot(rf_model)
```

The first plot shows the **error rate** as the forest grows. The **overall error** (black line) reflects the model’s total misclassification rate across both classes. The **class 0 error** (green) is the error rate specifically for non-cancellations, while the **class 1 error** (red) is the error rate for cancellations. The curves stabilize early, showing that the model converges well before 500 trees. Notably, class 1 error is lower than class 0 error, meaning the model is better at identifying **cancellations** than **non-cancellations**, which is common when the signal for cancellations is stronger.

The second figure shows the model’s **variable importance** using two metrics.  
- **MeanDecreaseAccuracy** measures how much predictive accuracy drops when each variable is permuted. Large drops indicate variables the model truly relies on.  
- **MeanDecreaseGini** measures how much a variable reduces impurity across all tree splits. This shows its structural importance within the forest.

Across both measures, the most influential features are **lead_time**, **average price per room**, and **number of special requests**, followed by several timing-related variables such as **arrival_month**, **arrival_date**, and **number of weekend nights**. Together, these results highlight that **booking timing, price sensitivity, and guest behavior signals** are the strongest predictors of cancellation risk in the dataset.


## Step 5: Predict
In our financial assumptions, a **false positive**—predicting a cancellation when the guest actually arrives—incurs a **\$500 walking cost**, which is **twice as expensive** as a false negative, where failing to predict a cancellation results in only **\$250 of lost revenue**. Because the cost of being wrong on a “cancel” prediction is substantially higher, we want to be **more conservative** when predicting cancellations. Therefore, using the default threshold of 0.5 would lead to too many false positives and unnecessary walking costs. To mitigate this, we adopt a **higher decision threshold** (0.70), ensuring we only predict a cancellation when the model is sufficiently confident. This approach minimizes expensive false positives while still capturing most true cancellations.

### Step 5.1: Predict LogReg
#### Basic LogReg
```{r, cache=TRUE}
logreg_prob <- predict(logreg_model, newdata = hotel_test, type = "response")
pred_logreg <- ifelse(logreg_prob >= 0.7, 1, 0)
```

#### Enhanced LogReg
```{r, cache=TRUE}
logreg_enh_prob <- predict(logreg_model_enhanced, newdata = hotel_test, type = "response")
pred_logreg_enh <- ifelse(logreg_enh_prob >= 0.7, 1, 0)
```

### Step 5.2: Predict KNN Model
Let's figure out the optimal k value. We want to balance specificity and accuracy effectively.
```{r, cache=TRUE}
results <- data.frame(k = numeric(), Accuracy = numeric(), Specificity = numeric())

for (k in seq(1, 31, by = 2)) {
  
  knn_pred <- knn(
    train = train_features,
    test  = test_features,
    cl    = train_labels,
    k     = k,
    prob  = TRUE
  )

  knn_prob <- ifelse(
    knn_pred == "1",
    attr(knn_pred, "prob"),   
    1 - attr(knn_pred, "prob") 
  )
  
  knn_binary <- ifelse(knn_prob >= 0.7, 1, 0)
  
  cm <- confusionMatrix(
    as.factor(knn_binary),
    as.factor(test_labels),
    positive = "1"
  )
  
  results <- rbind(
    results,
    data.frame(
      k = k,
      Accuracy = cm$overall["Accuracy"],
      Specificity = cm$byClass["Specificity"]
    )
  )
}

results

```
Based on our KNN tuning results, the optimal value of **k = 13** achieves the best balance for our business objective. Although its overall accuracy is **80.8%**, it delivers the **highest specificity (96.3%)** of all tested values. Because false positives (walking guests) carry the largest financial penalty in our cost structure, this high specificity makes k = 13 the most practical choice for minimizing expensive overbooking errors while still maintaining reasonable predictive performance.


Let's build our final model now.
```{r, cache=TRUE}
knn_final <- knn(
  train = train_features,
  test  = test_features,
  cl    = train_labels,
  k     = 13,
  prob  = TRUE
)

knn_prob_final <- ifelse(
  knn_final == "1",
  attr(knn_final, "prob"),
  1 - attr(knn_final, "prob")
)

knn_binary_final <- ifelse(knn_prob_final >= 0.7, 1, 0)
```


### Step 5.3: Predict ANN
```{r, cache=TRUE}
ann_prob_raw <- predict(ann_model, hotel_test_scaled)
ann_prob <- ann_prob_raw[, 1] 
pred_ann <- ifelse(ann_prob >= 0.7, 1, 0)
```

### Step 5.4: Predict Decision Tree
```{r, cache=TRUE}
dt_prob <- predict(m_decision, hotel_test, type = "prob")[, "1"]
pred_dt <- ifelse(dt_prob >= 0.7, 1, 0)
```

### Step 5.5: Predict SVM

#### Basic SVM
```{r, cache=TRUE}
svm_rbf_prob <- predict(hotel_svm_basic, hotel_test_scaled, type = "probabilities")
svm_rbf_prob_class1 <- svm_rbf_prob[, "1"]
pred_svm_basic <- ifelse(svm_rbf_prob_class1 >= 0.7, 1, 0)

```

#### Enhanced SVM
```{r, cache=TRUE}
svm_rbf_enhanced <- predict(hotel_svm_enhanced, hotel_test_scaled, type = "probabilities")
svm_rbf_enhanced_class1 <- svm_rbf_enhanced[, "1"]
pred_svm_enhanced <- ifelse(svm_rbf_enhanced_class1 >= 0.7, 1, 0)
```


### Step 5.6: Predict Random Forest
```{r, cache=TRUE}
rf_prob <- predict(rf_model, hotel_test, type = "prob")[, "1"]
pred_rf <- ifelse(rf_prob >= 0.7, 1, 0)
```


## Step 6: Evaluate

### Step 6.1: Evaluate LogReg Models
#### Basic LogReg
```{r}
confusionMatrix(as.factor(pred_logreg), as.factor(hotel_test$booking_status), positive = "1")
```
The baseline logistic regression model achieved an accuracy of **77.4%**, meaning it correctly classified a little over three out of every four reservations. However, its performance is highly unbalanced across classes. The model’s **sensitivity is only 39.7%**, meaning it detects fewer than half of all true cancellations. In contrast, its **specificity is very high at 95.9%**, indicating that it is excellent at correctly identifying guests who will *not* cancel.

**Business Case Usefulness: Low**

#### Enhanced LogReg
```{r}
confusionMatrix(as.factor(pred_logreg_enh), as.factor(hotel_test$booking_status), positive = "1")
```
The enhanced logistic regression improves slightly across all metrics, with better **sensitivity (42.3%)** and **balanced accuracy (69.4%).** However, it still misses the majority of true cancellations, limiting its usefulness for overbooking optimization. Although it reduces financial risk relative to the basic model, its predictive power is still not strong enough to reliably inform high-stakes overbooking decisions.

**Business Case Usefulness: Low–Moderate**

### Step 6.2: Evaluate KNN Model
```{r}
confusionMatrix(as.factor(knn_binary_final), as.factor(test_labels), positive = "1")
```

The KNN model shows a meaningful improvement with **accuracy at 80.9%** and **balanced accuracy at 72.8%.** **Sensitivity rises to 49.2%**, which means it detects about half of cancellations—a notable improvement over logistic regression. **Specificity remains high (96.3%)**, which helps avoid costly false positives. While KNN is not the strongest model available, it delivers moderate predictive capability and could be applied cautiously to guide overbooking under conservative thresholds.

**Business Case Usefulness: Moderate**

### Step 6.3: Evaluate ANN Model
```{r}
confusionMatrix(as.factor(pred_ann), as.factor(hotel_test_scaled$booking_status), positive = "1")
```
The ANN model achieves a modest balance between **sensitivity (41.9%)** and **specificity (95.7%)**. Its sensitivity is around the same mark as the linear models, and it still misses most true cancellations, limiting its usefulness for overbooking decisions where identifying high-risk reservations is critical. Specificity remains very strong, meaning the ANN reliably recognizes guests who will *not* cancel, but this comes at the cost of under-detecting cancellations. With a **balanced accuracy of 68.8%**, the ANN provides only a minor improvement in signal quality over simpler models.

**Business Case Usefulness: Moderate**



### Step 6.4: Evaluate Decision Tree
```{r}
confusionMatrix(as.factor(pred_dt), as.factor(hotel_test$booking_status), positive = "1")
```
The decision tree delivers strong performance across key metrics: **high sensitivity (68.4%), high specificity (96%), and leading balanced accuracy (82.2%).** This combination makes it effective at identifying both cancellations and non-cancellations. It minimizes both lost revenue and overbooking penalties, making it a very strong candidate for operational use. Its interpretability also helps hotel managers understand risk factors.

**Business Case Usefulness: High**


### Step 6.5: Evaluate SVM
#### Basic SVM
```{r}
confusionMatrix(as.factor(pred_svm_basic), as.factor(hotel_test_scaled$booking_status), positive = "1")
```
The basic SVM model achieves solid **sensitivity (54.6%) and balanced accuracy (75.1%)**, outperforming logistic regression and KNN. This means it captures a larger share of cancellations while maintaining good specificity. The model presents a good compromise between detecting cancellation risk and avoiding false positives.

**Business Case Usefulness: Moderate–High**


#### Enhanced SVM
```{r}
confusionMatrix(as.factor(pred_svm_enhanced), as.factor(hotel_test_scaled$booking_status), positive = "1")
```
The enhanced SVM model shows slightly higher overall accuracy (**83.6%**) and maintains a strong balance between sensitivity (**59.1%**) and specificity (**95.5%**). Its sensitivity improves over the basic SVM, allowing it to detect more true cancellations, while specificity remains essentially unchanged, continuing to limit costly false overbookings. Overall, it remains a robust and reliable model for supporting overbooking policies and is the stronger option compared to the basic SVM.

**Business Case Usefulness: Moderate–High**


### Step 6.6: Evaluate Random Forest
```{r}
confusionMatrix(as.factor(pred_rf), as.factor(hotel_test$booking_status), positive = "1")
```
The random forest is one of the strongest models in the portfolio, with excellent **accuracy (87.3%), high sensitivity (66.7%), very high specificity (97.4%), and strong balanced accuracy (82%).** This model effectively identifies cancellations while avoiding false alarms, giving hotels a powerful tool for making profitable overbooking decisions. Its strong predictive signal and superior error balance make it the best-performing model for the business use case.

**Business Case Usefulness: High**


### Summary Table

| Model                      | Accuracy | Kappa  | Sensitivity | Specificity | Balanced Accuracy |
|---------------------------|---------:|-------:|------------:|------------:|------------------:|
| LogReg (Basic)            | 0.7742 | 0.4098 | 0.3966 | 0.9588 | 0.6777 |
| LogReg (Enhanced)         | 0.7864 | 0.4443 | 0.4232 | 0.9640 | 0.6936 |
| KNN (Final, threshold 0.7)| 0.8085 | 0.5116 | 0.4915 | 0.9634 | 0.7275 |
| ANN                       | 0.7802 | 0.4299 | 0.4191 | 0.9567 | 0.6879 |
| Decision Tree             | 0.8688 | 0.6838 | **0.6838** | 0.9593 | **0.8215** |
| SVM (Basic)               | 0.8216 | 0.5540 | 0.5464 | 0.9561 | 0.7513 |
| SVM (Enhanced)            | 0.8355 | 0.5943 | 0.5914 | 0.9549 | 0.7731 |
| Random Forest             | **0.8732** | **0.6905** | 0.6670 | **0.9741** | 0.8205 |


Overall, the **Random Forest model is the best option** so far for hotels to identify cancellations and make profitable overbooking decisions. Alternatively, the Decision Tree model is also effective and may be incorporated into hotel operations.


## Step 7: Build Second Level Model Dataset

Now, we will build second level models by using the first level models. 

```{r}
stacked_data <- data.frame(
  logreg = logreg_enh_prob,
  knn = knn_prob_final,
  ann = ann_prob,
  decision_tree = dt_prob,
  svm = svm_rbf_enhanced_class1,
  random_forest = rf_prob,
  booking_status = hotel_test$booking_status
)

head(stacked_data)
```

## Step 8: Split Stacked Data
```{r}
trainprop <- 0.7
trainrows <- sample(1:nrow(stacked_data), trainprop * nrow(stacked_data))

stack_train <- stacked_data[trainrows, ]
stack_test  <- stacked_data[-trainrows, ]
```

## Step 9: Build Second Level Models

Before training the second-level stacked models, it is worth noting that our first-level models already achieve strong predictive performance.This suggests that a cost matrix is not strictly required to reach high accuracy or balanced error rates. However, because our goal is not just predictive performance but **business-aligned decision-making**, we incorporate thresholds and cost weighting to reflect the real financial tradeoffs faced by hotels. In our business case, a false positive (incorrectly predicting a guest will not cancel) results in an expensive overbooking penalty, while a false negative simply leads to an unused room. Therefore, we construct a cost matrix where **false positives are treated as twice as costly as false negatives**, aligning the stacked models with the operational costs of hotel overbooking. For our non-cost-sensitive models, we continue to utilize our threshold of 0.7.


### Step 9.1: Build Decision Tree without Cost Matrix
```{r}
model_stack_nocost <- C5.0(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status)
)
plot(model_stack_nocost)

```

### Step 9.2: Build Decision Tree with Cost Matrix
 

```{r}
cost_matrix <- matrix(c(0, 500, 250, 0), nrow = 2)
cost_matrix

model_stack_cost <- C5.0(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status),
  costs = cost_matrix
)
plot(model_stack_cost)

```

### Step 9.3: Build Random Forest without Class Weights
```{r}
rf_stack_nocost <- randomForest(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status),
  ntree = 500,
  mtry = sqrt(ncol(stack_train) - 1)
)
varImpPlot(rf_stack_nocost, main = "Variable Importance — Stacked RF (No Cost)")

```

### Step 9.4: Build Random Forest with Class Weights
```{r}
rf_stack <- randomForest(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status),
  ntree = 500,
  mtry = sqrt(ncol(stack_train) - 1),
  classwt = c("0" = 500, "1" = 250)
)
varImpPlot(rf_stack, main = "Variable Importance — Stacked RF (Cost Weighted)")

```

### Analysis
Across all four stacked models,the ensembles consistently place the most weight on the strongest base learners. In every model, the **Random Forest prediction** is the dominant input: it forms the first split in both stacked decision trees and shows the highest variable importance in both stacked random forests. The **Decision Tree** model is the next most influential, regularly appearing in upper-level splits and receiving the second-highest importance scores.

Notably, the stacked models also reveal that the **SVM model provides meaningful predictive signal**. In both the cost-weighted and non-cost-weighted versions of the stacked random forest, SVM ranks as the third most important contributor. It also appears repeatedly in mid-level splits of the stacked decision trees, indicating that it adds useful complementary information that the tree-based learners rely on when refining predictions.

KNN, logistic regression, and the ANN model contribute smaller amounts of information, generally appearing only in lower-level branches or receiving lower importance scores. They still add incremental value, but their influence on the ensemble is clearly weaker compared to Random Forest, Decision Tree, and SVM.

Overall, the stacked models naturally learn to trust the **top three base models—Random Forest, Decision Tree, and SVM—most heavily**, which aligns with their relatively strong standalone performance. This confirms that the stacking procedure is effectively capturing and combining the predictive strengths of the best individual learners.


## Step 10: Predict and Evaluate

### Step 10.1: Predict/Evaluate Decision Tree without Cost Matrix

```{r}
dt_no_cost_prob <- predict(model_stack_nocost, stack_test, type = "prob")[, "1"]
dt_no_cost_pred <- ifelse(dt_no_cost_prob >= 0.7, 1, 0)
confusionMatrix(
  as.factor(dt_no_cost_pred),
  as.factor(stack_test$booking_status),
  positive = "1"
)
```
This stacked decision tree without a cost matrix performs exceptionally well across all evaluation metrics. It achieves an accuracy of **89.1%**, a strong sensitivity of **74.3%**, and an excellent specificity of **96.4%**, meaning it correctly identifies most cancellations while keeping false overbookings very low. Its **balanced accuracy of 85.3%** is among the highest of all first and second-level models, demonstrating that the ensemble generalizes well across both classes. This strong combination of high sensitivity and high specificity makes the model especially effective for identifying high-risk reservations while avoiding costly overbooking penalties.

**Business Case Usefulness: High**


### Step 10.2: Predict/Evaluate Decision Tree with Cost Matrix
```{r}
dt_cost <- predict(model_stack_cost, stack_test, type = "class")

confusionMatrix(
  as.factor(dt_cost),
  as.factor(stack_test$booking_status),
  positive = "1"
)

```
When a cost matrix is applied to the stacked decision tree, the model becomes more conservative in its predictions. Sensitivity decreases to **70.4%**, meaning it identifies slightly fewer cancellations, but specificity increases to an exceptional **97.1%**, sharply reducing costly false positives. This behavior aligns with the business objective encoded in the cost matrix: avoiding overbookings (false positives) is twice as important as missing a cancellation (false negative).

Despite being more conservative, the model still delivers strong overall performance, achieving a high accuracy of **88.2%** and a robust balanced accuracy of **83.7%**. This makes the cost-weighted stacked decision tree a reliable option when hotel management prioritizes minimizing the financial risk of overbooking, even if it means missing a few additional cancellations.

**Business Case Usefulness: Moderate–High**



### Step 10.3: Predict/Evaluate Random Forest without Costs
```{r}
rf_stack_nocost_prob <- predict(rf_stack_nocost, stack_test, type = "prob")[, "1"]
rf_stack_nocost_pred <- ifelse(rf_stack_nocost_prob >= 0.7, 1, 0)

confusionMatrix(
  as.factor(rf_stack_nocost_pred),
  as.factor(stack_test$booking_status),
  positive = "1"
)
```
The stacked random forest without a cost matrix performs extremely well under the 0.7 threshold. It achieves a strong sensitivity of **71.4%**, meaning it identifies a large share of true cancellations, while maintaining an exceptionally high specificity of **97.4%**, one of the highest among all first- and second-level models. This balance minimizes costly false positives while still capturing most high-risk bookings. Its **balanced accuracy of 84.4%** reflects robust performance across both classes, even with the more conservative threshold.

**Business Case Usefulness: Moderate–High**


### Step 10.4: Predict/Evaluate Random Forest with Costs
```{r}
rf_stack_prob <- predict(rf_stack, stack_test, type = "prob")[, "1"]
rf_stack_pred <- ifelse(rf_stack_prob >= 0.5, 1, 0)

confusionMatrix(
  as.factor(rf_stack_pred),
  as.factor(stack_test$booking_status),
  positive = "1"
)

```
The stacked random forest with a cost matrix is one of the strongest performers across all models. With excellent sensitivity (**81.5%**) and strong specificity (**94.3%**), it captures a large majority of true cancellations while still keeping false positives relatively low. Its overall accuracy of **90.0%** is the highest among all first- and second-level models, and its **balanced accuracy of 87.9%** is also near the top, demonstrating outstanding performance across both classes.

**Business Case Usefulness: High**




### Summary Table (Including Stacked Models)

| Model                               | Accuracy | Kappa  | Sensitivity | Specificity | Balanced Accuracy |
|-------------------------------------|---------:|-------:|------------:|------------:|------------------:|
| LogReg (Basic)                      | 0.7742   | 0.4098 | 0.3966      | 0.9588      | 0.6777            |
| LogReg (Enhanced)                   | 0.7864   | 0.4443 | 0.4232      | 0.9640      | 0.6936            |
| KNN (Final, threshold 0.7)          | 0.8085   | 0.5116 | 0.4915      | 0.9634      | 0.7275            |
| ANN                                 | 0.7802   | 0.4299 | 0.4191      | 0.9567      | 0.6879            |
| Decision Tree (Base)                | 0.8688   | 0.6838 | 0.6838      | 0.9593      | 0.8215            |
| SVM (Basic)                         | 0.8216   | 0.5540 | 0.5464      | 0.9561      | 0.7513            |
| SVM (Enhanced)                      | 0.8355   | 0.5943 | 0.5914      | 0.9549      | 0.7731            |
| Random Forest (Base)                | 0.8732   | 0.6905 | 0.6670      | 0.9741      | 0.8205            |
| **Stacked DT (No Cost, thr = 0.7)** | 0.8905 | 0.7410 | 0.7429      | 0.9637      | 0.8533        |
| Stacked DT (Cost Matrix)            | 0.8820   | 0.7170 | 0.7036      | 0.9706      | 0.8371            |
| Stacked RF (No Cost, thr = 0.7)     | 0.8881   | 0.7317 | 0.7141      | **0.9744**      | 0.8443            |
| **Stacked RF (Cost Weighted)**      | **0.9004** | **0.7713** | **0.8150**      | 0.9428      | **0.8789**        |
           |
Overall, **the stacked random forest with the cost matrix is the best model** for hotel operational use.


## Step 11: Profit
Based on performance, we have determined that the stacked random forest with the cost matrix is most optimal. Next, we will confirm this by checking how each model affects hotels' financial performances. 

### Confusion Matrix 

| Case | Model Prediction | Actual Outcome | Financial Impact |
|------|--------------|----------------|-----------------|
| **True Positive (TP)** | "Will Cancel" | Guest cancels | **+ \$250** (revenue from replacement guest) |
| **True Negative (TN)** | "Will Not Cancel" | Guest does not cancel | **\$0** (no change in revenue) |
| **False Negative (FN)** | "Will Not Cancel" | Guest cancels | **– \$250** (lost booking revenue) |
| **False Positive (FP)** | "Will Cancel" | Guest shows up | **– \$500** (overbooking penalty) |

Profit Formula: 250 * TP - 250 * FN - 500 * FP

### Assumptions Behind the Financial Model

To keep the analysis consistent and interpretable, we base our financial calculations on the following assumptions:

- A walked guest always costs the hotel **$500**, and a canceled booking always loses **$250** unless resold.
- The hotel can reliably resell rooms predicted as cancellations (i.e., demand is sufficiently high).
- All reservations are treated as independent (no group bookings or corporate contracts).
- Customer mix, pricing, and seasonality remain similar to the historical data.
- The financial cost of guest dissatisfaction is already incorporated into the $500 walk penalty.

These assumptions reflect typical hotel operations but should be reviewed and adjusted when deploying the model in practice.


### Baseline Comparison (Before Model)
If the hotel does **no overbooking**, every canceled booking represents a loss of \$250.  
There were **3,553 total cancellations (TP + FN)**, resulting in  
**Baseline Profit = 3,553 × (–\$250) = –\$888,250**


### Profit Comparison Table — Base Models

| Model              | TN     | FP   | FN     | TP     | Profit Calculation                                                            | Raw Profit    |
|--------------------|--------|------|--------|--------|-------------------------------------------------------------------------------|---------------|
| LogReg (Basic)     | 11681  | 502  | 3593   | 2362   | 2362×250 – 502×500 – 3593×250 = 590,500 – 251,000 – 898,250                   | **–558,750**  |
| LogReg (Enhanced)  | 11744  | 439  | 3435   | 2520   | 2520×250 – 439×500 – 3435×250 = 630,000 – 219,500 – 858,750                   | **–448,250**  |
| KNN (0.7)          | 11737  | 446  | 3028   | 2927   | 2927×250 – 446×500 – 3028×250 = 731,750 – 223,000 – 757,000                   | **–248,250**  |
| ANN                | 11655  | 528  | 3459   | 2496   | 2496×250 – 528×500 – 3459×250 = 624,000 – 264,000 – 864,750                   | **–504,750**  |
| Decision Tree      | 11687  | 496  | 1883   | 4072   | 4072×250 – 496×500 – 1883×250 = 1,018,000 – 248,000 – 470,750                 | **+299,250**  |
| SVM (Basic)        | 11648  | 535  | 2701   | 3254   | 3254×250 – 535×500 – 2701×250 = 813,500 – 267,500 – 675,250                   | **–129,250**  |
| SVM (Enhanced)     | 11633  | 550  | 2433   | 3522   | 3522×250 – 550×500 – 2433×250 = 880,500 – 275,000 – 608,250                   | **–2,750**    |
| Random Forest      | 11867  | 316  | 1983   | 3972   | 3972×250 – 316×500 – 1983×250 = 993,000 – 158,000 – 495,750                   | **+339,250**  |


### Profit Comparison Table — Stacked Models
Scaled because the stacked test set is 30% of the original test set. 

| Model                       | TN   | FP  | FN  | TP   | Profit Calculation                                                     | Raw Profit | Scaled Profit |
|-----------------------------|------|-----|------|------|--------------------------------------------------------------------------|------------|----------------|
| Stacked DT (No Cost)        | 3505 | 132 | 464  | 1341 | 1341×250 – 132×500 – 464×250 = 335,250 – 66,000 – 116,000               | +153,250   | **+510,833**   |
| Stacked DT (Cost Matrix)    | 3530 | 107 | 535  | 1270 | 1270×250 – 107×500 – 535×250 = 317,500 – 53,500 – 133,750               | +130,250   | **+434,167**   |
| Stacked RF (No Cost)        | 3544 | 93  | 516  | 1289 | 1289×250 – 93×500 – 516×250 = 322,250 – 46,500 – 129,000                | +146,750   | **+489,167**   |
| Stacked RF (Cost Weighted)  | 3429 | 208 | 334  | 1471 | 1471×250 – 208×500 – 334×250 = 367,750 – 104,000 – 83,500              | +180,250   | **+600,833**   |

The stacked models dramatically outperform all first-level models in terms of profitability. While only two base models (Decision Tree and Random Forest) produced positive profit, **all four stacked models generate substantial financial gains**, with the **Cost-Weighted Stacked Random Forest delivering the highest projected profit (+600,833)**. In our case, stacking not only improves predictive accuracy but also translates directly into significantly better financial outcomes for hotel overbooking decisions.


### Final Recommendation:
Our final recommendation is the Cost-Weighted Stacked Random Forest. The weighted Random Forest is more stable, incorporates our financial costs directly into training, and better avoids the expensive false positives, making it the more reliable model for real hotel operations. It also outperforms all other models on metrics and profit. 

### How Much Is the Model Worth?

- **Baseline (Doing Nothing):**  
  5,955 × (–\$250) = **–\$1,488,750**

- **Best Model Profit:**  
  **+\$318,333**

- **Model Value:**  
  \$600,833 – (–\$1,488,750) = **\$2,089,583**

**The model is worth approximately $2.09 million.**

#### Net Savings Per Customer

We divide the total value of the model by the number of customers in the full test set:

- **Total customers:** 18,138  
- **Model value:** $2,089,583  
- **Net savings per customer:**  
  $2,089,583 ÷ 18,138 = **$115.20 per customer**

**The model generates approximately $115.20 in net financial savings for every customer booking.**


### Operational Interpretation (What This Means for a Real Hotel)

From an operational standpoint, the model can be embedded directly into a hotel’s reservation or revenue-management system to update overbooking levels in real time. By sharply reducing **false positives** (the most expensive error type), the model minimizes the risk of walking guests while still recovering revenue from likely cancellations. Even moderate improvements in predictive accuracy translate into over **$2.0 million in annualized financial value** in our dataset, making this an extremely impactful decision-support tool for high-volume hotels. 

However, it is important to recognize that the model relies on historical cancellation behavior, which may shift due to seasonality, economic factors, or changes in booking policies. Additionally, rebooking revenue and walk costs are hotel-specific and may differ from the fixed values assumed here, meaning periodic retraining and threshold tuning will be necessary to keep the model aligned with real-world conditions.


### Practical Scenarios: How Ron Can Use the Model

Now that Ron has a cancellation-prediction model, he can make smarter, data-driven decisions in daily operations. Here are several realistic examples of how the model improves decision-making:

**Scenario 1: Extra Demand on a Busy Night**  
Ron has 20 confirmed reservations for Saturday night and receives 10 more booking requests.  
The model identifies that **6 of the original guests have a high cancellation probability** based on the stacked Random Forest output.  
→ Ron confidently accepts 6 additional bookings, knowing he is unlikely to exceed capacity.  
→ Result: No walks, no empty rooms, and an estimated **+\$1,500** in recovered revenue.

**Scenario 2: Corporate Group Blocks 5 Rooms**  
A corporate client wants to hold 5 unpaid rooms for an event.  
The model classifies the group’s booking pattern as **high cancellation risk**, based on similar historical cases.  
→ Ron can safely oversell by 1–2 rooms to protect revenue.  
→ Result: Better utilization and reduced chance of unused inventory.

**Scenario 3: High-Risk Overbooking During a Holiday**  
On a holiday weekend, demand is extremely high.  
The model predicts that **most current reservations are low-risk (likely to show)**.  
→ Ron chooses *not* to overbook, avoiding expensive \$500 guest-walk penalties.  
→ Result: No angry customers, no compensation costs, smoother operations.

**Scenario 4: End-of-Month Revenue Boost**  
Ron is behind on his monthly revenue target.  
The model highlights a cluster of bookings with **elevated cancellation probability**, giving Ron the confidence to slightly oversell and recapture otherwise lost revenue.  
→ Result: A safe, predictable month-end performance without harming guest satisfaction.

These scenarios show how the model transforms Ron’s overbooking strategy from guesswork into a measurable, financially optimized process—directly tied to the hotel’s economic priorities.
