---
title: "Final Project Draft of Code"
author: "Third Time's the Charm"
date: "2025-10-23"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 0: Why?


**Dataset Chosen:**  
Hotel Reservations (~35,000 rows)  
Predicts hotel booking cancellations and potential no-shows. The business use case focuses on **optimizing overbooking decisions** to maximize hotel occupancy and revenue.


### Context (from Kaggle)
> “The online hotel reservation channels have dramatically changed booking possibilities and customers’ behavior. A significant number of hotel reservations are called-off due to cancellations or no-shows. The typical reasons for cancellations include change of plans, scheduling conflicts, etc. This is often made easier by the option to do so free of charge or preferably at a low cost which is beneficial to hotel guests but it is a less desirable and possibly revenue-diminishing factor for hotels to deal with.”

### Business Use Case
We are building a model to predict which hotel reservations are most likely to cancel.  
Although cancellations often result from a change of plans or scheduling conflicts, they impose a direct financial cost on the hotel.  
To mitigate these losses and **maintain 100% occupancy**, the hotel can strategically **overbook** based on model predictions of cancellation likelihood.

**Main Question:**  
Can we predict which hotel reservations are most likely to be canceled, and use these predictions to make more profitable overbooking decisions?

**Why it’s answerable with this dataset:**  
- The dataset includes **booking-level data** for over 35 000 reservations with both categorical and numerical features (lead time, room type, guest history, etc.).  
- The dependent variable `booking_status` is binary (Canceled / Not Canceled), making it well-suited for **classification via logistic regression**.  
- With exploratory cleaning and modeling, we can estimate how much predictive signal exists and whether cancellation behavior is predictable enough to support overbooking decisions.


### Financial Assumptions
- **Revenue per booked room:** \$250 per night on average price in a city
(source: https://www.coohom.com/article/how-much-do-hotel-rooms-cost)
- **Cancellations:** Guests can cancel free of charge at any time; each cancellation results in a \$250 revenue loss (from the cost of the room)  
- **Overbooking cost:** If the hotel overbooks and cannot accommodate a guest (because no cancellations occur), it incurs \$500 per night because the cost of walking a guest is ~2x the room cost in re-accommodation, transportation, goodwill, and/or compensation costs.
(souce: https://ecornell-impact.cornell.edu/the-cheapest-and-best-approach-to-overbooking/)
- **Variable costs per room:** Considered sunk and excluded from analysis (the cost of turning a room is identical regardless of occupancy).  
- **No difference** between cancellation and no-show costs.


### Cost–Benefit Matrix

| Case | Model Prediction | Actual Outcome | Financial Impact |
|------|--------------|----------------|-----------------|
| **True Positive (TP)** | "Will Cancel" | Guest cancels | **+ \$250** (revenue from replacement guest) |
| **True Negative (TN)** | "Will Not Cancel" | Guest does not cancel | **\$0** (no change in revenue) |
| **False Negative (FN)** | "Will Not Cancel" | Guest cancels | **– \$250** (lost booking revenue) |
| **False Positive (FP)** | "Will Cancel" | Guest shows up | **– \$500** (overbooking penalty) |


### Objective
Develop a predictive model that helps hotels **balance the trade-off** between overbooking risk and lost revenue due to cancellations, thereby maximizing **expected net revenue** and improving operational decision-making.


### Step 0.5: Load Packages
```{r}
library(caret)
library(neuralnet)
library(class)
library(C50)
library(ggplot2)
library(randomForest)
library(kernlab)
```


## Step 1: Load Data
```{r}
hotel <- read.csv("Hotel Reservations.csv", stringsAsFactors = TRUE)
```

## Step 2: Clean Data
```{r}
str(hotel)
summary(hotel)

```
Let's remove booking id. Let's also convert our dependent variable (booking status) into 0s and 1s.
```{r}
hotel$Booking_ID = NULL
hotel$booking_status <- ifelse(hotel$booking_status == "Canceled", 1, 0)
str(hotel)
summary(hotel)
```

## Step 3: Split Data
We have 36000 data entries. In order to build a second level model, we are going to keep 30% for evaluation.
Let's dummify and split the data.
```{r}
hotel_dummies <- as.data.frame(model.matrix(~ . - booking_status - 1, data = hotel))
hotel_dummies$booking_status <- hotel$booking_status

trainprop <- 0.7
set.seed(12345)
trainrows <- sample(1:nrow(hotel_dummies), trainprop * nrow(hotel_dummies))

hotel_train <- hotel_dummies[trainrows, ]
hotel_test  <- hotel_dummies[-trainrows, ]
str(hotel_train)
summary(hotel_train)
```

Let's now keep a normalized copy of the data for some of our models.
```{r}
minmax <- function(x) {
  if (is.numeric(x)) {
    return((x - min(x)) / (max(x) - min(x)))
  } else {
    return(x)
  }
}

hotel_train_scaled <- as.data.frame(lapply(hotel_train, minmax))
hotel_test_scaled <- as.data.frame(lapply(hotel_test, minmax))
hotel_train_scaled$booking_status <- as.factor(hotel_train$booking_status)
hotel_test_scaled$booking_status  <- as.factor(hotel_test$booking_status)

str(hotel_train_scaled)
summary(hotel_train_scaled)
```


## Step 4: Build Models

### Step 4.1: Build LogReg Model

Our baseline model will be a **vanilla Logistic Regression (LogReg)**, which predicts the probability of a booking being canceled.  
This serves as the foundation for comparison with future models (e.g., neural networks, decision trees, etc.).
```{r, cache=TRUE}
logreg_model <- glm(booking_status ~ ., data = hotel_train, family = "binomial")
```

Let's also build an enhanced model. To improve the predictive performance of our baseline logistic regression model, we introduce several **interaction terms**. These interactions allow the model to capture relationships between variables that jointly influence cancellation behavior. Including these terms helps the model represent more complex customer patterns that cannot be captured by individual predictors alone.

#### **1. `lead_time × avg_price_per_room`**
This interaction captures how the impact of lead time changes depending on room price.
- Customers who book **high-priced rooms far in advance** may be more likely to cancel as plans, budgets, or alternatives change.
- Customers booking inexpensive rooms far in advance may behave differently.
This term allows the model to account for situations where **price modifies the effect of lead time** on cancellations.

#### **2. `no_of_special_requests × repeated_guest`**
This interaction differentiates behavior between repeat guests and first-time customers.
- Repeat guests with many special requests may be more committed (e.g., loyal or planned stays).
- First-time guests with many requests may be more likely to cancel if expectations shift.
This term helps the model identify how **customer loyalty interacts with booking complexity**.

#### **3. `no_of_weekend_nights × no_of_week_nights`**
This interaction captures the structure of the guest’s stay.
- Longer, more complex trips (spanning both weekend and weekday nights) may carry higher cancellation risks.
This term helps the model evaluate how the **overall length and nature of the stay** influence cancellation behavior.

#### **4. `arrival_month × market_segment_type` (applied across dummy variables)**
After dummification, `market_segment_type` becomes several binary indicators (e.g., Online, Offline, Corporate).  
Interacting `arrival_month` with each segment dummy allows the model to learn **segment-specific seasonal cancellation patterns**:
- Corporate travelers may cancel more during holiday months.
- Online/OTA bookings often surge—and cancel more—during peak travel seasons or promotional periods.
- Offline bookings (e.g., via agents) may follow entirely different seasonal trends.  
By including these interactions, the model captures how **seasonality modifies cancellation behavior across different booking channels**.
If you want, I can also generate a matching R code block (perfectly formatted) and a clean narrative paragraph describing the enhanced model for your deliverable.


```{r, cache=TRUE}
logreg_model_enhanced <- glm(
  booking_status ~ 
    . +
    lead_time * avg_price_per_room +
    no_of_special_requests * repeated_guest +
    no_of_weekend_nights * no_of_week_nights +
    arrival_month * market_segment_typeComplementary +
    arrival_month * market_segment_typeCorporate +
    arrival_month * market_segment_typeOffline +
    arrival_month * market_segment_typeOnline,
  data = hotel_train,
  family = "binomial"
)


```


### Step 4.2: Build KNN Model
Here, we separate our predictors and target variables into training and testing sets. We will predict and evaluate in step 5. 
```{r}
train_features <- subset(hotel_train_scaled, select = -booking_status)
train_labels <- hotel_train_scaled$booking_status
test_features <- subset(hotel_test_scaled, select = -booking_status)
test_labels <- hotel_test_scaled$booking_status
```

### Step 4.3: Build ANN Model
We are now building an ANN model with a hidden layer of 5 neurons. With structured data, typically a shallow network can capture enough valuable relationships. 
```{r, cache = TRUE}
ann_model <- neuralnet(
  booking_status ~ .,
  data = hotel_train_scaled,
  lifesign = "full",
  stepmax = 1e6,
  hidden = 5
)
plot(ann_model)
```


### Step 4.4: Build Decision Tree Model
We will now build a normal decision tree model. We will not impose a cost matrix here but we will in the second level decision tree. 
```{r, cache=TRUE}
m_decision <- C5.0(
  as.factor(booking_status) ~ .,
  data = hotel_train
)

plot(m_decision)
```
The decision tree relies most heavily on a few key predictors. **Lead time** is by far the strongest driver of cancellation behavior, appearing in virtually every major split. The model also heavily uses **number of special requests**, **repeated guest status**, and **number of week nights** to differentiate likely cancellations from reliable bookings. Additionally, **price-related features** (such as `avg_price_per_room`) and **market segment type** help the tree distinguish different types of guests. Overall, the tree prioritizes guest commitment indicators (lead time, repeat status), stay characteristics (week nights, weekend nights), and booking channel/price information to identify cancellation risk.

### Step 4.5: Build SVM
We will train a basic SVM with a RBF kernel.
```{r, cache=TRUE}
hotel_svm_basic <- ksvm(
  as.factor(booking_status) ~ .,
  data = hotel_train_scaled,
  kernel = "rbfdot",
  prob.model = TRUE
)
```

We will also build an enhanced SVM model. We  refine the model by tuning two key hyperparameters:

**C (Regularization parameter):**  
Controls the trade-off between maximizing the margin and minimizing classification errors.

- A **larger C** (e.g., 5) penalizes misclassifications more heavily, producing a **smaller margin** and a tighter, more detailed fit to the data (lower bias, higher variance).  
- A **smaller C** allows more errors but produces a **wider margin**, leading to a smoother, more generalizable model (higher bias, lower variance).

**σ (Sigma in the RBF kernel):**  
Determines how far the influence of each support vector extends.

- A **smaller σ** (e.g., 0.05) makes each support vector’s influence very local, enabling the model to capture **fine-grained, nonlinear boundaries** (risk of overfitting).  
- A **larger σ** smooths the decision boundary, reducing overfitting but potentially missing subtle structure in the data.

This improved SVM aims to better separate likely cancellations from non-cancellations by allowing a more flexible, finely tuned boundary.
```{r, cache=TRUE}
hotel_svm_enhanced <- ksvm(
  as.factor(booking_status) ~ .,
  data = hotel_train_scaled,
  kernel = "rbfdot",
  C = 1,
  kpar = list(sigma = 0.1),
  prob.model = TRUE
)
```



### Step 4.6: Build Random Forest
We now train a Random Forest classifier using the unscaled dataset. We include 500 trees and use the square-root rule for mtry, which is standard for classification tasks.
```{r, cache=TRUE}
hotel_train$booking_status <- as.factor(hotel_train$booking_status)
hotel_test$booking_status  <- as.factor(hotel_test$booking_status)

rf_model <- randomForest(
  x = hotel_train[, -which(names(hotel_train) == "booking_status")],
  y = hotel_train$booking_status,
  ntree = 500,
  mtry = sqrt(ncol(hotel_train) - 1),
  importance = TRUE
)

plot(rf_model)
varImpPlot(rf_model)
```
**Note about what the RF finds important**

## Step 5: Predict
In our financial assumptions, a **false positive**—predicting a cancellation when the guest actually arrives—incurs a **\$500 walking cost**, which is **twice as expensive** as a false negative, where failing to predict a cancellation results in only **\$250 of lost revenue**. Because the cost of being wrong on a “cancel” prediction is substantially higher, we want to be **more conservative** when predicting cancellations. Therefore, using the default threshold of 0.5 would lead to too many false positives and unnecessary walking costs. To mitigate this, we adopt a **higher decision threshold** (0.70), ensuring we only predict a cancellation when the model is sufficiently confident. This approach minimizes expensive false positives while still capturing most true cancellations.

### Step 5.1: Predict LogReg
```{r, cache=TRUE}
logreg_prob <- predict(logreg_model, newdata = hotel_test, type = "response")
pred_logreg <- ifelse(logreg_prob >= 0.7, 1, 0)

logreg_enh_prob <- predict(logreg_model_enhanced, newdata = hotel_test, type = "response")
pred_logreg_enh <- ifelse(logreg_enh_prob >= 0.7, 1, 0)
```

### Step 5.2: Predict KNN Model
Let's figure out the optimal k value. We want to balance specificity and accuracy effectively.
```{r, cache=TRUE}
results <- data.frame(k = numeric(), Accuracy = numeric(), Specificity = numeric())

for (k in seq(1, 31, by = 2)) {
  
  knn_pred <- knn(
    train = train_features,
    test  = test_features,
    cl    = train_labels,
    k     = k,
    prob  = TRUE
  )

  knn_prob <- ifelse(
    knn_pred == "1",
    attr(knn_pred, "prob"),   
    1 - attr(knn_pred, "prob") 
  )
  
  knn_binary <- ifelse(knn_prob >= 0.7, 1, 0)
  
  cm <- confusionMatrix(
    as.factor(knn_binary),
    as.factor(test_labels),
    positive = "1"
  )
  
  results <- rbind(
    results,
    data.frame(
      k = k,
      Accuracy = cm$overall["Accuracy"],
      Specificity = cm$byClass["Specificity"]
    )
  )
}

results

```
Based on our KNN tuning results, the optimal value of **k = 13** achieves the best balance for our business objective. Although its overall accuracy is **81.9%**, it delivers the **highest specificity (96.1%)** of all tested values. Because false positives (walking guests) carry the largest financial penalty in our cost structure, this high specificity makes k = 13 the most practical choice for minimizing expensive overbooking errors while still maintaining reasonable predictive performance.


Let's build our final model now.
```{r, cache=TRUE}
knn_final <- knn(
  train = train_features,
  test  = test_features,
  cl    = train_labels,
  k     = 13,
  prob  = TRUE
)

knn_prob_final <- ifelse(
  knn_final == "1",
  attr(knn_final, "prob"),
  1 - attr(knn_final, "prob")
)

knn_binary_final <- ifelse(knn_prob_final >= 0.7, 1, 0)
```


### Step 5.3: Predict ANN
```{r, cache=TRUE}
ann_prob <- predict(ann_model, hotel_test_scaled)
pred_ann <- ifelse(ann_prob >= 0.7, 1, 0)
```

### Step 5.4: Predict Decision Tree
```{r, cache=TRUE}
dt_prob <- predict(m_decision, hotel_test, type = "prob")[, "1"]
pred_dt <- ifelse(dt_prob >= 0.7, 1, 0)
```

### Step 5.5: Predict SVM

Basic SVM
```{r, cache=TRUE}
svm_rbf_prob <- predict(hotel_svm_basic, hotel_test_scaled, type = "probabilities")
svm_rbf_prob_class1 <- svm_rbf_prob[, "1"]
pred_svm_basic <- ifelse(svm_rbf_prob_class1 >= 0.7, 1, 0)

```

Enhanced SVM
```{r, cache=TRUE}
svm_rbf_enhanced <- predict(hotel_svm_enhanced, hotel_test_scaled, type = "probabilities")
svm_rbf_enhanced_class1 <- svm_rbf_enhanced[, "1"]
pred_svm_enhanced <- ifelse(svm_rbf_enhanced_class1 >= 0.7, 1, 0)
```


### Step 5.6: Predict Random Forest
```{r, cache=TRUE}
rf_prob <- predict(rf_model, hotel_test, type = "prob")[, "1"]
pred_rf <- ifelse(rf_prob >= 0.7, 1, 0)
```


## Step 6: Evaluate

### Step 6.1: Evaluate LogReg Models
Basic LogReg
```{r}
confusionMatrix(as.factor(pred_logreg), as.factor(hotel_test$booking_status), positive = "1")
```
The baseline logistic regression model achieved an accuracy of **77.7%**, meaning it correctly classified a little over three out of every four reservations. However, its performance is highly unbalanced across classes. The model’s **sensitivity is only 40.6%**, meaning it detects fewer than half of all true cancellations. In contrast, its **specificity is very high at 95.6%**, indicating that it is excellent at correctly identifying guests who will *not* cancel.


Enhanced LogReg
```{r}
confusionMatrix(as.factor(pred_logreg_enh), as.factor(hotel_test$booking_status), positive = "1")
```
Very similar, but we will choose enhanced. Better Accuracy, sensitivity, specificity, kappa, balanced accuracy. 

### Step 6.2: Evaluate KNN Model
```{r}
confusionMatrix(as.factor(knn_binary_final), as.factor(test_labels), positive = "1")
```

### Step 6.3: Evaluate ANN Model
```{r}
confusionMatrix(as.factor(pred_ann), as.factor(hotel_test_scaled$booking_status), positive = "1")
```

### Step 6.4: Evaluate Decision Tree
```{r}
confusionMatrix(as.factor(pred_dt), as.factor(hotel_test$booking_status), positive = "1")
```

### Step 6.5: Evaluate SVM
Basic SVM
```{r}
confusionMatrix(as.factor(pred_svm_basic), as.factor(hotel_test_scaled$booking_status), positive = "1")
```

Enhanced SVM
```{r}
confusionMatrix(as.factor(pred_svm_enhanced), as.factor(hotel_test_scaled$booking_status), positive = "1")
```
Enhanced better. better accuracy, kappa, specificity. 

### Step 6.6: Evaluate Random Forest
```{r}
confusionMatrix(as.factor(pred_rf), as.factor(hotel_test$booking_status), positive = "1")
```

### Summary Table

| Model                      | Accuracy | Kappa  | Sensitivity | Specificity | Balanced Accuracy |
|---------------------------|---------:|-------:|------------:|------------:|------------------:|
| LogReg (Basic)            | 0.7768 | 0.4167 | 0.4061 | 0.9565 | 0.6813 |
| LogReg (Enhanced)         | 0.7886 | 0.4490 | 0.4289 | 0.9629 | 0.6959 |
| KNN (Final, threshold 0.7)| 0.8185 | 0.5405 | 0.5243 | 0.9611 | 0.7427 |
| ANN                       | 0.8088 | 0.5389 | 0.5964 | 0.9117 | 0.7541 |
| Decision Tree             | 0.8797 | 0.7154 | **0.7399** | 0.9475 | **0.8437** |
| SVM (Basic)               | 0.8117 | 0.5517 | 0.6243 | 0.9026 | 0.7634 |
| SVM (Enhanced)            | 0.8194 | 0.5604 | 0.5961 | 0.9276 | 0.7618 |
| Random Forest             | **0.8854** | **0.7232** | 0.7104 | **0.9703** | 0.8403 |


RF and DT stand out from the rest.


## Step 7: Build Second Level Model Dataset
```{r}
stacked_data <- data.frame(
  logreg = logreg_enh_prob,
  knn = knn_prob_final,
  ann = ann_prob,
  decision_tree = dt_prob,
  svm = svm_rbf_enhanced_class1,
  random_forest = rf_prob,
  booking_status = hotel_test$booking_status
)

head(stacked_data)
```

## Step 8: Split Stacked Data
```{r}
trainprop <- 0.7
trainrows <- sample(1:nrow(stacked_data), trainprop * nrow(stacked_data))

stack_train <- stacked_data[trainrows, ]
stack_test  <- stacked_data[-trainrows, ]
```

## Step 9: Build Second Level Models

### Step 9.1: Build Decision Tree with Cost Matrix
```{r}
model_stack_nocost <- C5.0(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status)
)
```


### Step 9.2: Build Decision Tree with Cost Matrix
```{r}
cost_matrix <- matrix(c(0, 500, 250, 0), nrow = 2)
cost_matrix
```
Build Decision Tree
```{r}
model_stack_cost <- C5.0(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status),
  costs = cost_matrix
)
```
### Step 9.3: Build Random Forest without Class Weights
```{r}
rf_stack_nocost <- randomForest(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status),
  ntree = 500,
  mtry = sqrt(ncol(stack_train) - 1)
)
```


### Step 9.4: Build Random Forest with Class Weights
```{r}
rf_stack <- randomForest(
  x = stack_train[, -which(colnames(stack_train) == "booking_status")],
  y = as.factor(stack_train$booking_status),
  ntree = 500,
  mtry = sqrt(ncol(stack_train) - 1),
  classwt = c("0" = 500, "1" = 250)
)


```


## Step 10: Predict and Evaluate
### Step 10.1: Predict/Evaluate Decision Tree without Cost Matrix
```{r}
dt_no_cost_prob <- predict(model_stack_nocost, stack_test, type = "prob")[, "1"]
dt_no_cost_pred <- ifelse(dt_no_cost_prob >= 0.7, 1, 0)
confusionMatrix(
  as.factor(dt_no_cost_pred),
  as.factor(stack_test$booking_status),
  positive = "1"
)
```


### Step 10.2: Predict/Evaluate Decision Tree with Cost Matrix
```{r}
dt_cost <- predict(model_stack_cost, stack_test, type = "class")

confusionMatrix(
  as.factor(dt_cost),
  as.factor(stack_test$booking_status),
  positive = "1"
)

```

### Step 10.3: Predict/Evaluate Random Forest without Costs
We threshold at 0.7.
```{r}
rf_stack_nocost_prob <- predict(rf_stack_nocost, stack_test, type = "prob")[, "1"]
rf_stack_nocost_pred <- ifelse(rf_stack_nocost_prob >= 0.7, 1, 0)

confusionMatrix(
  as.factor(rf_stack_nocost_pred),
  as.factor(stack_test$booking_status),
  positive = "1"
)
```


### Step 10.4: Predict/Evaluate Random Forest with Costs
Since costs are already incorporated, we threshold at 0.5.
```{r}
rf_stack_prob <- predict(rf_stack, stack_test, type = "prob")[, "1"]
rf_stack_pred <- ifelse(rf_stack_prob >= 0.5, 1, 0)

confusionMatrix(
  as.factor(rf_stack_pred),
  as.factor(stack_test$booking_status),
  positive = "1"
)

```

### Summary Table (Including Stacked Models)

| Model                               | Accuracy | Kappa  | Sensitivity | Specificity | Balanced Accuracy |
|-------------------------------------|---------:|-------:|------------:|------------:|------------------:|
| LogReg (Basic)                      | 0.7768   | 0.4167 | 0.4061      | 0.9565      | 0.6813            |
| LogReg (Enhanced)                   | 0.7886   | 0.4490 | 0.4289      | 0.9629      | 0.6959            |
| KNN (Final, threshold 0.7)          | 0.8185   | 0.5405 | 0.5243      | 0.9611      | 0.7427            |
| ANN                                 | 0.8088   | 0.5389 | 0.5964      | 0.9117      | 0.7541            |
| Decision Tree (Base)                | 0.8797   | 0.7154 | 0.7399      | 0.9475      | 0.8437            |
| SVM (Basic)                         | 0.8117   | 0.5517 | 0.6243      | 0.9026      | 0.7634            |
| SVM (Enhanced)                      | 0.8194   | 0.5604 | 0.5961      | 0.9276      | 0.7618            |
| Random Forest (Base)                | 0.8854   | 0.7232 | 0.7104      | 0.9703      | 0.8403            |
| Stacked DT (No Cost, thr = 0.7)     | **0.8949**   | **0.7579** | **0.8110**      | 0.9358      | **0.8734**            |
| Stacked DT (Cost Matrix)            | 0.8873   | 0.7323 | 0.7418      | 0.9581      | 0.8500            |
| Stacked RF (No Cost, thr = 0.7)     | 0.8882   | 0.7324 | 0.7287      | **0.9658**      | 0.8473            |
| Stacked RF (Cost Weighted)          | **0.8949*   | 0.7577 | 0.8092      | 0.9367      | 0.8729            |



## Step 11: Profit



### Confusion Matrix 

| Case | Model Prediction | Actual Outcome | Financial Impact |
|------|--------------|----------------|-----------------|
| **True Positive (TP)** | "Will Cancel" | Guest cancels | **+ \$250** (revenue from replacement guest) |
| **True Negative (TN)** | "Will Not Cancel" | Guest does not cancel | **\$0** (no change in revenue) |
| **False Negative (FN)** | "Will Not Cancel" | Guest cancels | **– \$250** (lost booking revenue) |
| **False Positive (FP)** | "Will Cancel" | Guest shows up | **– \$500** (overbooking penalty) |

Profit Formula: 250 * TP - 250 * FN - 500 * FP

### Baseline Comparison (Before Model)
If the hotel does **no overbooking**, every canceled booking represents a loss of \$300.  
There were **3,553 total cancellations (TP + FN)**, resulting in  
**Baseline Profit = 3,553 × (–\$250) = –\$888,250**


### Profit Comparison Table — Base Models

| Model                   | TN   | FP  | FN   | TP   | Profit Calculation                                                    | Raw Profit |
|-------------------------|------|-----|------|------|-----------------------------------------------------------------------|------------|
| LogReg (Basic)          | 7011 | 319 | 2110 | 1443 | 1443×250 – 319×500 – 2110×250 = 360,750 – 159,500 – 527,500           | –326,250   |
| LogReg (Enhanced)       | 7058 | 272 | 2029 | 1524 | 1524×250 – 272×500 – 2029×250 = 381,000 – 136,000 – 507,250           | –262,250   |
| KNN (0.7)               | 7045 | 285 | 1690 | 1863 | 1863×250 – 285×500 – 1690×250 = 465,750 – 142,500 – 422,500           | –99,250    |
| ANN                     | 6683 | 647 | 1434 | 2119 | 2119×250 – 647×500 – 1434×250 = 529,750 – 323,500 – 358,500           | –152,250   |
| Decision Tree (Base)    | 6945 | 385 | 924  | 2629 | 2629×250 – 385×500 – 924×250 = 657,250 – 192,500 – 231,000            | +233,750   |
| SVM (Basic)             | 6616 | 714 | 1335 | 2218 | 2218×250 – 714×500 – 1335×250 = 554,500 – 357,000 – 333,750           | –136,250   |
| SVM (Enhanced)          | 6799 | 531 | 1435 | 2118 | 2118×250 – 531×500 – 1435×250 = 529,500 – 265,500 – 358,750           | –94,750    |
| Random Forest (Base)    | 7112 | 218 | 1029 | 2524 | 2524×250 – 218×500 – 1029×250 = 631,000 – 109,000 – 257,250           | +264,750   |

### Profit Comparison Table — Stacked Models
Scaled because the stacked test set is 30% of the original test set. 

| Model                       | TN   | FP  | FN  | TP  | Profit Calculation                                  | Raw Profit | Scaled Profit |
|-----------------------------|------|-----|-----|-----|-----------------------------------------------------|-------|------------|
| Stacked DT (No Cost)        | 2055 | 141 | 202 | 867 | 867×250 – 141×500 – 202×250 = 216,750 – 70,500 – 50,500     | +95,750 | +318,333 |
| Stacked DT (Cost Matrix)    | 2104 | 92  | 276 | 793 | 793×250 – 92×500 – 276×250 = 198,250 – 46,000 – 69,000  | +83,250    | +277,500   |
| Stacked RF (No Cost)        | 2121 | 75  | 290 | 779 | 779×250 – 75×500 – 290×250 = 194,750 – 37,500 – 72,500      | +84,750   | +282,500 |
| Stacked RF (Cost Weighted)  | 2057 | 139 | 204 | 865 | 865×250 – 139×500 – 204×250 = 216,250 – 69,500 – 51,000    | +95,750    | +318,333 |

### Final Recommendation:
Our final recommendation is the Cost-Weighted Stacked Random Forest. While the no-cost stacked decision tree performed similarly, the weighted Random Forest is more stable, incorporates our financial costs directly into training, and better avoids the expensive false positives, making it the more reliable model for real hotel operations.

### How Much Is the Model Worth?

- **Baseline (Doing Nothing):**  
  3,553 × (–\$250) = **–\$888,250**

- **Best Model Profit:**  
  **+\$318,333**

- **Model Value:**  
  \$318,333 – (–\$888,250) = **\$1,206,583**

**The model is worth approximately \$1.21 million.**


