---
title: "Intermediate Deliverable 1"
author: "Third Time's the Charm"
date: "2025-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 0: Why?

**Dataset Chosen:**  
Hotel Reservations (~35,000 rows)  
Predicts hotel booking cancellations and potential no-shows. The business use case focuses on **optimizing overbooking decisions** to maximize hotel occupancy and revenue.


### Context (from Kaggle)
> “The online hotel reservation channels have dramatically changed booking possibilities and customers’ behavior. A significant number of hotel reservations are called-off due to cancellations or no-shows. The typical reasons for cancellations include change of plans, scheduling conflicts, etc. This is often made easier by the option to do so free of charge or preferably at a low cost which is beneficial to hotel guests but it is a less desirable and possibly revenue-diminishing factor for hotels to deal with.”

### Business Use Case
We are building a model to predict which hotel reservations are most likely to cancel.  
Although cancellations often result from a change of plans or scheduling conflicts, they impose a direct financial cost on the hotel.  
To mitigate these losses and **maintain 100% occupancy**, the hotel can strategically **overbook** based on model predictions of cancellation likelihood.

**Main Question:**  
Can we predict which hotel reservations are most likely to be canceled, and use these predictions to make more profitable overbooking decisions?

**Why it’s answerable with this dataset:**  
- The dataset includes **booking-level data** for over 35 000 reservations with both categorical and numerical features (lead time, room type, guest history, etc.).  
- The dependent variable `booking_status` is binary (Canceled / Not Canceled), making it well-suited for **classification via logistic regression**.  
- With exploratory cleaning and modeling, we can estimate how much predictive signal exists and whether cancellation behavior is predictable enough to support overbooking decisions.


### Financial Assumptions
- **Revenue per booked room:** \$300 per night  
- **Cancellations:** Guests can cancel free of charge at any time; each cancellation results in a \$300 revenue loss.  
- **Overbooking cost:** If the hotel overbooks and cannot accommodate a guest (because no cancellations occur), it incurs \$400 per night in re-accommodation, goodwill, or compensation costs.  
- **Variable costs per room:** Considered sunk and excluded from analysis (the cost of turning a room is identical regardless of occupancy).  
- **No difference** between cancellation and no-show costs.


### Cost–Benefit Matrix

| Case | Model Prediction | Actual Outcome | Financial Impact |
|------|--------------|----------------|-----------------|
| **True Positive (TP)** | "Will Cancel" | Guest cancels | **+ \$300** (revenue from replacement guest) |
| **True Negative (TN)** | "Will Not Cancel" | Guest does not cancel | **\$0** (no change in revenue) |
| **False Negative (FN)** | "Will Not Cancel" | Guest cancels | **– \$300** (lost booking revenue) |
| **False Positive (FP)** | "Will Cancel" | Guest shows up | **– \$400** (overbooking penalty) |


### Objective
Develop a predictive model that helps hotels **balance the trade-off** between overbooking risk and lost revenue due to cancellations, thereby maximizing **expected net revenue** and improving operational decision-making.


### Step 0.5: Load Packages
```{r}
library(caret)
```


## Step 1: Load Data
```{r}
hotel <- read.csv("Hotel Reservations.csv", stringsAsFactors = TRUE)
```

## Step 2: Clean Data
```{r}
str(hotel)
summary(hotel)

```
Let's remove booking id. Let's also convert our dependent variable (booking status) into 0s and 1s.
```{r}
hotel$Booking_ID = NULL
hotel$booking_status <- ifelse(hotel$booking_status == "Canceled", 1, 0)
str(hotel)
summary(hotel)
```

## Step 3: Split Data
We have about 36000 data entries. 15% of that is close to 5500 entries which should be more than comprehensive for a test set. 
```{r}
trainprop <- 0.85
set.seed(12345)
trainrows <- sample(1:nrow(hotel), trainprop*nrow(hotel))

hotel_train <- hotel[trainrows, ]
hotel_test <- hotel[-trainrows, ]
```


## Step 4: Build LogReg model

Our baseline model will be a **vanilla Logistic Regression (LogReg)**, which predicts the probability of a booking being canceled.  
This serves as the foundation for comparison with future models (e.g., neural networks, decision trees, etc.).
```{r}
logreg_model <- glm(booking_status ~ ., data = hotel_train, family = "binomial")
```


## Step 5: Predict
Let's use a threshold of 0.5 for now.
```{r}
logreg_predictions <- predict(logreg_model, newdata = hotel_test, type = "response")
logreg_pred_class <- ifelse(logreg_predictions >= 0.5, 1, 0)
```

## Step 6: Evaluate
```{r}
confusionMatrix(as.factor(logreg_pred_class), as.factor(hotel_test$booking_status), positive = "1")
```
The baseline logistic regression model achieved an **accuracy of 81.5%**, meaning it correctly classified roughly four out of five reservations.  
The model’s **sensitivity (65%)** indicates that it successfully identified about two-thirds of actual cancellations, while its **specificity (89%)** shows strong performance in recognizing non-cancellations.  
This balance suggests the model is slightly conservative—it’s better at avoiding false alarms (predicting cancellations that don’t happen) than at catching every cancellation.  
For a hotel aiming to minimize overbooking costs, this trade-off is acceptable for out vanilla model, though future tuning will hopefully allow us to improve recall in order to maximize profits. 


## Step 7: Profit
Using the financial assumptions from **Step 0**, we can estimate the model’s impact on revenue compared to doing nothing (i.e., no overbooking policy).

### Confusion Matrix Summary
| Outcome | Count | Description |
|------|------|-------------|
| **True Positives (TP)** | 1,153 | Predicted “will cancel” and guest did cancel |
| **True Negatives (TN)** | 3,281 | Predicted “will not cancel” and guest did not cancel |
| **False Positives (FP)** | 393 | Predicted “will cancel” but guest showed up |
| **False Negatives (FN)** | 615 | Predicted “will not cancel” but guest canceled |

### Financial Impact
| Case | Profit / Loss per Case | Count | Total Impact |
|------|------------|--------|---------|
| **TP** | +\$300 | 1,153 | +\$345,900 |
| **TN** | \$0 | 3,281 | \$0 |
| **FP** | –\$400 | 393 | –\$157,200 |
| **FN** | –\$300 | 615 | –\$184,500 |

**Total Expected Profit = \$345,900 – \$157,200 – \$184,500 = +\$4,200**

### Baseline Comparison (Before Model)
If the hotel does **no overbooking**, every canceled booking represents a loss of \$300.  
There were **1,768 total cancellations (TP + FN)**, resulting in  
**Baseline Profit = 1,768 × (–\$300) = –\$530,400**

### Interpretation
- **Without the model:** expected loss of \$530,400  
- **With the model:** expected gain of \$4,200  
- an overall **improvement of about \$534,600** in expected revenue on the test sample. 

Even this simple logistic regression adds meaningful value by identifying likely cancellations and allowing profitable overbooking.  
Future models with higher recall could further reduce lost revenue and increase profitability.

